---
title: "Survey state plot reprex"
author: "Brian McGreal"
date: "2026-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggpattern)
library(here)
library(MARSS)
library(panelr)
library(readxl)
library(tidyverse)
here::i_am("code/primary/09-TScomparison.R")
```

```{r loaddata_hidden, message=FALSE, warning=FALSE, echo=FALSE}
ssm_coho <- readRDS(file=here::here("data", "clean", "ssm_coho_Afull.rds"))
load(file=here::here("data", "clean", "nosa_cohoPOP.Rda"))
key_coho <- read_excel(here("data", "clean", "xtT_statekey.xlsx"), sheet = "coho")
load(file=here::here("data", "clean", "populations_list.Rda"))
pop_list <- pop_list |> 
  filter(CommonPopName !="Lostine River Spring Chinook")
pop_list$COMMONPOPNAME[pop_list$ESAPOPNAME == "Salmon, Chinook (Upper Willamette River ESU) Clackamas River - spring"] <- "Clackamas River - Spring"
pop_list$COMMONPOPNAME[pop_list$ESAPOPNAME == "Salmon, Chinook (Lower Columbia River ESU) Clackamas River - fall"] <- "Clackamas River - Fall"
pop_list$COMMONPOPNAME[pop_list$ESAPOPNAME == "Salmon, Chinook (Lower Columbia River ESU) Sandy River - late fall"] <- "Sandy River - Fall"
pop_list$COMMONPOPNAME[pop_list$ESAPOPNAME == "Salmon, Chinook (Lower Columbia River ESU) Sandy River - spring"] <- "Sandy River - Spring"
pop_names <- pop_list[-c(1, 3, 4, 6:10)]
keyPM_coho <- read_excel(here("data", "clean", "ytT_statekey.xlsx"), sheet = "coho")
```

## R Markdown

Here is how I'm coding my population-level plots of observed time series and underlying states along with their confidence intervals. First I pull in the MARSS model object ssm_coho, as well as two keys (keyX and keyY) which relate the model output to the proper underlying population ID.

```{r loaddata_visible}
# load data
ssm_coho <- readRDS(file=here::here("data", "clean", "ssm_coho_Afull.rds"))
keyX_coho <- read_excel(here("data", "clean", "xtT_statekey.xlsx"), sheet = "coho")
keyY_coho <- read_excel(here("data", "clean", "ytT_statekey.xlsx"), sheet = "coho")
```

Next I extract from the MARSS object state estimates and confidence intervals, as well as Y_t predictions and those intervals as well. This is where I'm concerned I might not be grabbing the right information from the fitted model.

```{r extract}
# pull fitted states from MARSS object
states_coho <- tsSmooth(ssm_coho, type = "xtT")

predobs_coho <- predict(ssm_coho, type = "ytT", interval = "prediction", level = 0.95)
obs_coho <- predobs_coho$pred
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
